---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Elbird

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/Elbird)](https://CRAN.R-project.org/package=Elbird)
[![](https://cranlogs.r-pkg.org/badges/Elbird)](https://cran.r-project.org/package=Elbird)
<!-- badges: end -->

The goal of Elbird is to provide R Wrapper functions in [kiwipiepy](https://github.com/bab2min/kiwipiepy) package.

## Installation

### Pre required

You need to install conda before installing Elbird.

``` r
install.packages("reticulate")
reticulate::install_miniconda()
```

### CRAN [*NOT YET*]
You can install the released version of Elbird from [CRAN](https://CRAN.R-project.org) with:

```r
install.packages("Elbird")
```

### Github
You can install the development version of Elbird from [github](https://github.com/mrchypark/Elbird) with:

``` r
install.packages("remotes")
remotes::install_github("mrchypark/Elbird")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(Elbird)
analyze_tbl("안녕하세요 kiwi 형태소 분석기의 R wrapper인 Elbird를 소개합니다.")
analyze_tbl(c("새롭게 작성된 패키지 입니다.", "tidytext와의 호환을 염두하고 작성하였습니다."))
analyze_tidytext("안녕하세요 kiwi 형태소 분석기의 R wrapper인 Elbird를 소개합니다.")
analyze_tt(c("새롭게 작성된 패키지 입니다.", "tidytext와의 호환을 염두하고 작성하였습니다."))

```


### With tidytext

```{r}
suppressMessages(library(dplyr))
library(stringr)
library(tidytext)
library(presidentSpeechKr)

spidx %>% 
  filter(president == "이명박") %>% 
  filter(str_detect(title, "취임사")) %>% 
  pull(link) %>% 
  get_speech(paragraph = T) %>%
  select(paragraph, content) -> tar
tar
tar %>% 
  unnest_tokens(
    input = content,
    output = word,
    token = analyze_tt
    )
```

## tag

Tag list that used in [kiwipiepy](https://github.com/bab2min/kiwipiepy) package.

<table>
<tr><th>대분류</th><th>태그</th><th>설명</th></tr>
<tr><th rowspan='5'>체언(N)</th><td>NNG</td><td>일반 명사</td></tr>
<tr><td>NNP</td><td>고유 명사</td></tr>
<tr><td>NNB</td><td>의존 명사</td></tr>
<tr><td>NR</td><td>수사</td></tr>
<tr><td>NP</td><td>대명사</td></tr>
<tr><th rowspan='5'>용언(V)</th><td>VV</td><td>동사</td></tr>
<tr><td>VA</td><td>형용사</td></tr>
<tr><td>VX</td><td>보조 용언</td></tr>
<tr><td>VCP</td><td>긍정 지시사(이다)</td></tr>
<tr><td>VCN</td><td>부정 지시사(아니다)</td></tr>
<tr><th rowspan='1'>관형사</th><td>MM</td><td>관형사</td></tr>
<tr><th rowspan='2'>부사(MA)</th><td>MAG</td><td>일반 부사</td></tr>
<tr><td>MAJ</td><td>접속 부사</td></tr>
<tr><th rowspan='1'>감탄사</th><td>IC</td><td>감탄사</td></tr>
<tr><th rowspan='9'>조사(J)</th><td>JKS</td><td>주격 조사</td></tr>
<tr><td>JKC</td><td>보격 조사</td></tr>
<tr><td>JKG</td><td>관형격 조사</td></tr>
<tr><td>JKO</td><td>목적격 조사</td></tr>
<tr><td>JKB</td><td>부사격 조사</td></tr>
<tr><td>JKV</td><td>호격 조사</td></tr>
<tr><td>JKQ</td><td>인용격 조사</td></tr>
<tr><td>JX</td><td>보조사</td></tr>
<tr><td>JC</td><td>접속 조사</td></tr>
<tr><th rowspan='5'>어미(E)</th><td>EP</td><td>선어말 어미</td></tr>
<tr><td>EF</td><td>종결 어미</td></tr>
<tr><td>EC</td><td>연결 어미</td></tr>
<tr><td>ETN</td><td>명사형 전성 어미</td></tr>
<tr><td>ETM</td><td>관형형 전성 어미</td></tr>
<tr><th rowspan='1'>접두사</th><td>XPN</td><td>체언 접두사</td></tr>
<tr><th rowspan='3'>접미사(XS)</th><td>XSN</td><td>명사 파생 접미사</td></tr>
<tr><td>XSV</td><td>동사 파생 접미사</td></tr>
<tr><td>XSA</td><td>형용사 파생 접미사</td></tr>
<tr><th rowspan='1'>어근</th><td>XR</td><td>어근</td></tr>
<tr><th rowspan='9'>부호, 외국어, 특수문자(S)</th><td>SF</td><td>종결 부호(. ! ?)</td></tr>
<tr><td>SP</td><td>구분 부호(, / : ;)</td></tr>
<tr><td>SS</td><td>인용 부호 및 괄호(' " ( ) [ ] < > { } ― ‘ ’ “ ” ≪ ≫ 등)</td></tr>
<tr><td>SE</td><td>줄임표(…)</td></tr>
<tr><td>SO</td><td>붙임표(- ~)</td></tr>
<tr><td>SW</td><td>기타 특수 문자</td></tr>
<tr><td>SL</td><td>알파벳(A-Z a-z)</td></tr>
<tr><td>SH</td><td>한자</td></tr>
<tr><td>SN</td><td>숫자(0-9)</td></tr>
<tr><th rowspan='1'>분석 불능</th><td>UN</td><td>분석 불능<sup>*</sup></td></tr>
<tr><th rowspan='4'>웹(W)</th><td>W_URL</td><td>URL 주소<sup>*</sup></td></tr>
<tr><td>W_EMAIL</td><td>이메일 주소<sup>*</sup></td></tr>
<tr><td>W_HASHTAG</td><td>해시태그(#abcd)<sup>*</sup></td></tr>
<tr><td>W_MENTION</td><td>멘션(@abcd)<sup>*</sup></td></tr>
</table>

<sup>*</sup> 세종 품사 태그와 다른 독자적인 태그입니다.
